---
title: "Section 1. General Trend Analysis"
format:
  docx:
    toc: true
    number-sections: true
    fig-dpi: 300
  html:
    toc: true
    toc-float: true
    code-fold: true
editor: visual
execute:
  echo: false
  warning: false
  message: false
params:
  pkg_path:     null
  spec_file:    null
  results_file: null
  min_n:        10
---

```{r setup}
# pkg_path is the absolute path to the package root, passed by the driver
# script. This is required because the QMD may render in a temp directory
# where here::here() cannot locate the DESCRIPTION file.
pkg_root <- params$pkg_path

# Load package functions.
# If pkg_root contains a source-level DESCRIPTION + R/ directory (dev mode),
# use devtools::load_all(). Otherwise the package is installed — use library().
if (!is.null(pkg_root) &&
    file.exists(file.path(pkg_root, "DESCRIPTION")) &&
    file.exists(file.path(pkg_root, "R"))) {
  devtools::load_all(pkg_root)
} else {
  library(ReportR)
}

# Source the assessment spec — puts assessment_spec in scope.
spec_path <- if (file.exists(params$spec_file)) {
  params$spec_file
} else {
  file.path(pkg_root, params$spec_file)
}
source(spec_path)

# Load pre-computed results from RDS.
general_trend_results <- readRDS(params$results_file)
min_n <- general_trend_results$params$min_n
```

## Introduction

```{r subtitle, results='asis'}
cat("###", assessment_spec$program$name, "\n")
```

This section presents general trend findings for `r assessment_spec$program$name`.
Results are organized into three areas: (1) scale score trends over time,
(2) proficiency trends, and (3) score distribution patterns. All analyses
exclude cells with fewer than `r min_n` students.

---

## Scale Score Trends

### Mean Scale Scores Over Time

```{r score-narrative, results='asis'}
cat(general_trend_results$score_trends$text$narrative)
```

```{r score-table}
tbl <- general_trend_results$score_trends$table

flextable::flextable(tbl) |>
  flextable::bold(part = "header") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::set_table_properties(layout = "autofit", width = 1)
```

```{r mean-trend-plot, fig.width=9, fig.height=4}
general_trend_results$score_trends$plot$mean_trend
```

```{r mean-sd-plot, fig.width=9, fig.height=4}
general_trend_results$score_trends$plot$mean_sd
```

### Grade-Level Score Profiles

```{r grade-profile-plot, fig.width=9, fig.height=4}
general_trend_results$score_trends$plot$grade_profile
```

```{r median-iqr-plot, fig.width=9, fig.height=4}
if (!is.null(general_trend_results$score_trends$plot$median_iqr)) {
  general_trend_results$score_trends$plot$median_iqr
}
```

---

## Proficiency Trends

### Percent Proficient Over Time

```{r prof-narrative, results='asis'}
cat(general_trend_results$proficiency$text$narrative)
```

```{r prof-table}
prof_tbl <- general_trend_results$proficiency$table

display_tbl <- prof_tbl |>
  dplyr::mutate(
    pct_proficient = round(pct_proficient, 1),
    GRADE = paste0("Grade ", GRADE)
  ) |>
  dplyr::arrange(SUBJECT, GRADE, YEAR)

flextable::flextable(display_tbl) |>
  flextable::bold(part = "header") |>
  flextable::fontsize(size = 9, part = "all") |>
  flextable::merge_v(j = c("SUBJECT", "GRADE")) |>
  flextable::set_table_properties(layout = "autofit", width = 1)
```

```{r prof-trend-plot, fig.width=9, fig.height=4}
general_trend_results$proficiency$plot$trend
```

### Proficiency Heatmap

```{r prof-heatmap-plot, fig.width=9, fig.height=4}
general_trend_results$proficiency$plot$heatmap
```

---

## Scale Score Distributions

### Distribution Shifts Over Time

```{r density-narrative, results='asis'}
cat(general_trend_results$distributions$text$narrative)
```

```{r density-plots, fig.width=10, fig.height=3, results='asis'}
density_plots <- general_trend_results$distributions$plot$density

for (subj in names(density_plots)) {
  p <- density_plots[[subj]]
  if (!is.null(p)) {
    print(p)
    cat("\n\n")
  }
}
```
